visualFields 1.x

### 2024-08-21 vignette.Rmd
Reaction to modifications by imarinfr.

To install the newestversion of the package, run:
```{r eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("imarinfr/vf1/source")
```

@@

# What is the visual field?
The visual field of an eye is the extent of the world that can be seen. It refers to both the central and the peripheral vision. It is measured in degrees of visual angle, with the center of vision being $0^\circ$, points on the temporal and superior regions of the visual field being positive angles and those in the inferior and nasal being negative. On healthy eyes, the sensitivity to light contrast changes as the magnitude of the visual angle increases (or as it is commonly said in perimetry, eccentricity). The difference in light sensitivity over the visual field as a function of eccentricity is known as the Traquair's *hill of vision* [@IMF:Traquair1938]. The figure below shows the hill of vision as rendered by Aulhorn & Harms [@IMF:Aul1967]. Notice in the discontinuity in the hill of vision in the lower right panel. Sensitivity in this region, centered at approximately $(15^\circ, -3^\circ)$ in the visual field is is completely blind, as it corresponds to the anatomical region of the retina known as the *blind spot*, where the optical fiber exits from the eye towards the visual cortex.

@@

# What is visual field testing?
Visual field testing measures a patient's visual sensitivity at different locations of the visual field. Its purpose is to reconstruct as best as possible within the confines and limitations of a clinical setting the *hill of vision*. The most common type of visual field testing is *static automated perimetry (SAP)*, which is performed with in a dome-shaped device, so that the distance between the stimulus and the eye equal for all locations within the visual field. [I WOULD DELETE THE FOLLOWING AS IT MAY CONFUSE RATHER THAN CLARIFY GIVEN THAT NOBODY USES CAMPIMETRY ANYMORE, AT LEAST IN THE UK OR US] In contrast, *campimetry* is performed with a flat screen.

@@

The visual fields package was created to analyze data from *SAP*. In SAP, the goal is to quantify sensitivity by increasing and decreasing the light differential (or contrast) between stimulus and background and asking the observer to press the button when they detect the presence of the visual stimulus. The contrast threshold is determined at each location of the visual field as the one for which the observer is able to detect the presence of the stimulus 50% of the times. Such thresholds are estimated at a fixed number of locations forming a regular grid as the one in the figure above. There is another common form of perimetry, *kinetic perimetry*, where a light spot is moved from the periphery towards the center along a number of meridians. In this form of perimetry the location where the spot is first detected by the subject is noted and those points at different meridians are connected to find isopters of equal sensitivity on the retina. The statistical analysis for this form of perimetry is very different than for static perimetry and beyond the scope of this visualFields package.

# Performing visual field tests
For the typical visual field test, an observer (usually a patient in an ophthalmology or optimetry clinic) is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld dome (or hemisphere). The perimeter is located in a dimly lit, calm room and the observer has been explained what to do by an experienced technician before the examination. In his hand, the observer has access to a clicker.

@@

At the end of the test, the inverse of the contrast threshold, $\Delta L_{\rm threshold}$ at each location is recorded in a decibel scale (dB). This dB scale is normally defined as attenuation from the maximum stimulus possible $\Delta L_{\rm max}$ as:
\begin{equation}
dB = -10\ \rm log \left (\frac{\Delta L_{\rm thr}}{\Delta L_{\rm max}} \right)\ .
\end{equation}
For the most common devices, The Humprhery Field Analyzer and the Octopus 900, the maximum luminance differential $\Delta L_{\rm max}$ is $10000\ (\rm apostilb) / \pi = 3183.1\ cd\ m^{-2}$.

@@

## Grid patterns
The spatial position of the points $l_1$ to $l_k$ [CAN WE USE K FOR LOCATIONS, TO PROTECT N FOR NUMBER OF VISUAL FIELDS OR SAMPLE SIZE?] in the visual field is found in the location map (locmap) and depends on the pattern used by the perimeter. When analyzing data, you need to specify the locmap.

@@

Locations of the locmap that cover the blind spot are reported in the locmaps lists under the variable bs. If the blind spot is outside the pattern (for example in the 10° visual fields) or is spared by the grid as in the Octopus G1 pattern, the list item bs is set to numeric(0).

@@

# visualFields as an aid for clinical assessment of the visual field
In clinical practice, visual field testing is used identify decreases in visual function caused by disease. Such *visual field defects* or *scotomas* can be caused by diseases that affect tissue anywhere along the optic pathway. This ranges from the optic media (cornea / lens / vitreous) which transmit light, to the retina, which detects lights and preprocesses visual information, to the optic nerve, which transmits information, and up to the visual cortex.

Often, clinicians can draw reliable conclusions about the localisation of the disease by the pattern of the visual field defects. For example, an infarction of the right visual cortex leads complete blindness in the left hemispheres of both eyes with a very distinct border along the vertical meridian. Therefore, the patient does not perceive anything to the right of the point they look at. On the other side of the vertical meridian the visual field is normal. In contrast, patients with glaucoma start with areas of slightly decreased light sensitivity often in the upper or lower hemifield with a bow-shaped (arcuate) pattern cause by the localisation of damage in the optic nerve head.

@@

The environment variable locini contains the number of the first column that contains the actual visual field date. Each column $l_{n}$ corresponds to one location in the visual field test. A concise way to select the columns $l_1$ to $l_{n}$ is to use the *getvfcols* function.

@@

There mustn't be any columns after these columns. Use the *vffilter* function to select specific fields and teh *vfisvalid* function to check whether a table fulfills the criteria. The latter will also tell you what the problem is when it doesn't.

@@

setdefaults()
###[I HAVE CHANGED THE EXAMPLE TO ONE FROM THE RETEST DATASET BECAUSE THE ONE FROM SUNY/IU IS TOO EXTREME]

@@

# Psychophysical testing
[THIS SECTION COULD BE MERGED WITH "Performing visual field tests"]
Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception. This is performed by varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses. In visual field testing, the observer usually presses a button when a light spot is perceived.

# Measurements
[THIS SECTION COULD BE AFTER "Performing visual field tests"]
Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response. It consists of the constant stimulus parameters (size, presentation time, background luminance), the stimulus location, the stimulus intensity and whether the subject pressed the button within a predefined time window to indicate that he saw the stimulus.

## False negatives and false positives
[THIS SECTION COULD BE AFTER "Performing visual field tests" TOO?]

# Mathematical derivatives of contrast thresholds for statisitcal analysis
## Identifying defective locations on the visual field
[A PART OF THIS HAS BEEN EXPLAINED IN MY EDITED VERSION OF "Performing visual field tests". I HAVE REWORKED THIS SECTION. SEE IF MY EDITS MAKE SENSE]
As mentioned earlier, the sensitivity in dB of attenuation from the maximum stimulus at each location is computed from the contrast thresholds.

@@

With these predictable and unpredictable differences in the pointwise sensitivity measurements, how can we provide tools and reports to aid the clinical assessment of the visual field? Or more to the point, how can we discriminate between healthy locations and locations that are damaged or suspected of being damaged? Clinically it is more useful to compute and report the deviations from an age-corrected mean normal visual field, since it becomes easier the identification of spatial patterns of damage thus facilitating the clinical assessment of the visual field.

@@

The resulting *deviations* from age-corrected mean normal visual field are also measured in dB and they are interpreted as follows. In the Humphrey Field Analyzer, a negative deviation value means that the sensitivity at that location is below mean normal. By contrast, on Octopus perimeters, negative is above and positive below mean normal. In this package, those differences are called *total deviation* and their interpretation are as for the Humphrey perimeter. In case you use visualFields with data from a different perimeter it is important to know what is the definition of *deviation* and change the sign if necessary.

@@

[CORD: I ADDED THIS TO SHOW THAT VISUALFIELDS IS NOT JUST TO ANALYZE THE DATA IN A CERTAIN WAY, YOU GENERATE YOUR OWN NORMATIVE VALUES AND WORK WITH THEM. SO FROM HERE ALL EXAMPLES ARE FROM nv INSTEAD OF getnv(). AT THE END OF THE DOCUMENT, WE CAN EXPLAIN setnv(), getnv(), setlocini(), setdefaults(), ETC. OK?]

@@

# Structure function correlations in glaucoma patients
In the last years, techniques that provide detailed imaging of the retina and the optic nerve have made incredible advances. Therefore, correlation between structure and function is becoming even more important than ever.

When [SINCE THIS IS MORE FOCUSED TO GLAUCOMA, I REMOVED PHOTORECEPTORS AND PUT INSTEAD GANGLION CELLS, AS PHOTORECEPTORS REMAIN INTACT IN GLAUCOMA] ganglion cells are lost in characteristic patterns at the retinal, [I DO NOT UNDERSTAND THIS SENTENCE] the scotoma always occurs in the directions in space that is perceived by that retinal area.

@@

@@

bundle_path <- data.frame(y = 4:90,
                          x_l19 = path_function_l19(4:90),
                          x_l23 = path_function_l23(4:90),
                          x_l47 = path_function_l47(4:90))
ggplot(bundle_path, aes(y = y)) +
  geom_line(aes(x = x_l19)) +
  geom_point(data = locations, aes(x = -psi, y = r, color = labels, shape = labels == "fixation")) +
  geom_line(aes(x = x_l23), color = "red") +
  geom_line(aes(x = x_l47), color = "blue") +
  coord_polar(start = pi / 2, direction = -1) +
  scale_x_continuous(breaks = c(-180, -90, 0, 90, 180), limits = c(-180, 180)) +
  scale_y_continuous(limits = c(0, 60)) +
  scale_color_manual(values = c("l19" = "black", "l23" = "red", "l47" = "blue", "fixation" = "black")) +
  guides(color = FALSE, shape = FALSE)

```
THERE ARE SOME WARNINGS IN THIS CODE THAT I DO NOT KNOW HOW TO TACKLE

@@

@@

# Using a different perimeter
Different manufacturers have different conventions in perimetry. For example, clinicians working with Humphrey Field Analyzers (Carl-Zeiss-Meditec) are often used to report loss of function as negative values (e.g., MD: mean *deviation* of $-10$ dB), while those working with Octopus perimeters (Haag-Streit) often report loss as a positive value (e.g., MD: mean *defect* of 10 dB). Furthermore, clinicians working with Octopus perimeters often use different grids, even though the 24-2 and 10-2 grids are also available. Usually, fields cannot be directly compared because other setting ($L_max$) are also different and because different algorithms are used.

@@

For a custom grid of location, it is possible to create and set the location map and its graphical parameters. visualFields has a function `vfgpar()` to generate graphical parameters for a given grid. Here is an example with the Octopus "G1" map, that uses `vfread` to read visualField's ready visual field data.
```{r}
# read the Octopus G1 coordinates and create the locmap
g1 <- list(coord = read.csv("./g1.csv"), name = "G1", desc = "The G1 location map used by Octopus perimeters", bs = numeric())
str(g1)
# set the G1 locmap
setlocmap(g1)
# create and set the graphical parameters
setgpar(vfgpar(g1$coord))
# read visual fields and plot the sensitivities
vfplot(vfread("./vfg1.csv"), type = "s")
```

>>>>

To install the newest version of the package, run:
```{r eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("imarinfr/vf1/source")
```

@@

# What is the visual field?
The visual field of an eye is the extent of the world that can be seen. It refers to both the central and the peripheral vision. It is measured in degrees of visual angle, with the center of vision being $0^\circ$, points on the temporal and superior regions of the visual field being positive angles and those in the inferior and nasal being negative. In healthy eyes, the sensitivity to light contrast decreases as the visual angle increases (or as it is commonly said in perimetry, eccentricity). The difference in light sensitivity over the visual field as a function of eccentricity is known as the Traquair's *hill of vision* [@IMF:Traquair1938]. The figure below shows the hill of vision as rendered by Aulhorn & Harms [@IMF:Aul1967]. Notice the discontinuity in the hill of vision in the lower right panel. Sensitivity in this region, the *blind spot*, centered at approximately $(15^\circ, -3^\circ)$ in the visual field corresponds to the optic nerve head [THE BLIND SPOT IS NOT AN ANATOMICAL TERM, BUT A PERIMETRIC ONE], where the optical fiber exits from the eye towards the visual cortex.

@@

# What is visual field testing?
Visual field testing measures a patient's visual sensitivity at different locations of the visual field. Its purpose is to reconstruct as best as possible within the confines and limitations of a clinical setting the *hill of vision*. The most common type of visual field testing is *static automated perimetry (SAP)*, which is performed with in a dome-shaped device, so that the distance between the stimulus and the eye equal for all locations within the visual field. In contrast, *campimetry* is performed with a flat screen. [CAMPIMETRY MAY BE USED MORE OFTEN IN THE FUTURE BECAUSE COMPUTER SCREENS/ VR DEVICES MAY BE USED FOR TELEMEDICINE; LEAVE IT?]

@@

The visual fields package was created to analyze data from *SAP*. In SAP, the goal is to quantify sensitivity by increasing and decreasing the light differential (or contrast) between stimulus and background and asking the observer to press the button when they detect the presence of the visual stimulus. The contrast threshold is determined at each location of the visual field as the one for which the observer is able to detect the presence of the stimulus 50% of the times. Such thresholds are estimated at a fixed number of locations forming a regular grid as the one in the figure above.

There is another common form of perimetry, *kinetic perimetry*, where a light spot is moved from the periphery towards the center along a number of meridians. In this form of perimetry the location where the spot is first detected by the subject is noted and those points at different meridians are connected to find isopters of equal sensitivity on the retina. The statistical analysis for this form of perimetry is very different than for static perimetry and beyond the scope of this visualFields package.

# Performing visual field tests
For the typical visual field test, an observer (usually a patient in an ophthalmology or optometry clinic) is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld dome (or hemisphere). The perimeter is located in a dimly lit, calm room and the observer has been explained what to do by an experienced technician before the examination. In his hand, the observer has access to a clicker.

@@

At the end of the test, the inverse of the contrast threshold, $\Delta L_{\rm threshold}$ at each location is recorded in a decibel scale (dB). This dB scale is normally defined as attenuation from the maximum stimulus possible $\Delta L_{\rm max}$ as:
\begin{equation}
dB = -10\ \rm log \left (\frac{\Delta L_{\rm thr}}{\Delta L_{\rm max}} \right)\ .
\end{equation}
For the most common devices, the Humprhery Field Analyzer and the Octopus 900, the maximum luminance differential $\Delta L_{\rm max}$ is $10000\ (\rm apostilb) / \pi = 3183.1\ cd\ m^{-2}$.

@@

## Psychophysical testing
Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception. This is performed by varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses. In visual field testing, the observer usually presses a button when a light spot is perceived.

## Grid patterns
The spatial position of the points $l_1$ to $l_k$ in the visual field is found in the location map (locmap) and depends on the pattern used by the perimeter. When analyzing data, you need to specify the locmap.

@@

Locations of the locmap that cover the blind spot are reported in the locmaps lists under the variable bs. If the blind spot is outside the pattern (for example in the 10° visual fields) or is spared by the grid as in the Octopus G1 pattern, the list item bs is set to `r numeric(0)`.

@@

## Measurements
Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response. It consists of the constant stimulus parameters (size, presentation time, background luminance), the stimulus location, the stimulus intensity and whether the subject pressed the button within a predefined time window to indicate that he saw the stimulus.

## False negatives and false positives

# Clinical applications of perimetry
In clinical practice, visual field testing is used to identify decreases in visual function caused by disease. Such *visual field defects* or *scotomas* can be caused by diseases that affect tissue anywhere along the optic pathway. This ranges from the optic media (cornea / lens / vitreous) which transmit light, to the retina, which detects lights and preprocesses visual information, to the optic nerve, which transmits information, and up to the visual cortex.

Often, clinicians can draw reliable conclusions about the localisation of the disease by the pattern of the visual field defects. For example, an infarction of the right visual cortex leads complete blindness in the left hemispheres of both eyes with a very distinct border along the vertical meridian. Therefore, the patient does not perceive anything to the right of the point he looks at. On the other side of the vertical meridian the visual field is normal. In contrast, patients with glaucoma start with areas of slightly decreased light sensitivity often in the upper or lower hemifield with a bow-shaped (arcuate) pattern cause by the localisation of damage in the optic nerve head.

@@

The environment variable locini contains the number of the first column that contains the actual visual field date. Each column $l_{x}$ corresponds to one of {k} locations in the visual field test. A concise way to select the columns $l_1$ to $l_{k}$ is to use the *getvfcols* function.

@@

There mustn't be any columns after these columns. Use the *vffilter* function to select specific fields and the *vfisvalid* function to check whether a table fulfills the criteria. The latter will also tell you what the problem is when it doesn't.

@@

@@

# Mathematical derivatives of contrast thresholds for statisitcal analysis

## Identifying defective locations on the visual field

As mentioned earlier, the sensitivity in dB at each location is calculated from the contrast threshold relative to the maximal contrast.

@@

With these predictable and unpredictable differences in the pointwise sensitivity measurements, how can we provide tools and reports to aid the clinical assessment of the visual field? Or more to the point, how can we discriminate between healthy locations and locations that are damaged or suspected of being damaged? Clinically, it is more useful to compute and report the deviations from an age-corrected mean normal visual field, since the identification of spatial patterns of damage becomes easier, thus facilitating the clinical assessment of the visual field.

@@

The resulting *deviations* from age-corrected mean normal visual field are also measured in dB and they are interpreted as follows. By convention, in the Humphrey Field Analyzer, a negative deviation value means that the sensitivity at that location is below mean normal. In contrast, on Octopus perimeters, negative is above and positive below mean normal (therefore, values are called *defects*). In this package, those differences are called *total deviation* and their interpretation are as for the Humphrey perimeter. In case you use visualFields with data from a different perimeter it is important to know what is the definition of *deviation* and change the sign if necessary.

@@

@@

# Structure function correlations in glaucoma patients
In the last years, techniques that provide detailed imaging of the retina and the optic nerve have made incredible advances. Therefore, correlation between structure and function is becoming more important than ever.

@@

```{r, warning=FALSE}

@@

bundle_path <- data.frame(y = 4:90,
                          x_l19 = path_function_l19(4:90),
                          x_l23 = path_function_l23(4:90),
                          x_l47 = path_function_l47(4:90))
ggplot(bundle_path, aes(y = y)) +
  geom_line(aes(x = x_l19)) +
  geom_point(data = locations, aes(x = -psi, y = r, color = labels, shape = labels == "fixation")) +
  geom_line(aes(x = x_l23), color = "red") +
  geom_line(aes(x = x_l47), color = "blue") +
  coord_polar(start = pi / 2, direction = -1) +
  scale_x_continuous(breaks = c(-180, -90, 0, 90, 180), limits = c(-180, 180)) +
  scale_y_continuous(limits = c(0, 60)) +
  scale_color_manual(values = c("l19" = "black", "l23" = "red", "l47" = "blue", "fixation" = "black")) +
  guides(color = "none", shape = "none")

@@

[IVAN: I THOUGHT THAT PSI WAS THE ANGLE OF INCIDENCE IN THE OPTIC NERVE HEAD. THEREFORE, I CHOSE A POLAR PLOT AND NOT AN X-Y-PLOT. ALSO, I THOUGHT THAT THE 0° ANGLE OF INCIDENCE WAS DEFINED AS THE LINE BETWEEN FOVEA AND ONH. I DID NOT TURN THE POLAR COORDINATED ACCORDINGLY EVEN THOUGHT I MENTIONED THAT THE FOVEA WAS "TILTED DOWNWARD". BUT I DO NOT HAVE MUCH EXPERIENCE WITH THE MODEL, SO PLEASE CORRECT ME IF I'M WRONG. HOW WOULD I CALCULATE XY-COORDINATES?]

@@

# Using a different perimeter
Different manufacturers have different conventions in perimetry. For example, clinicians working with Humphrey Field Analyzers (Carl-Zeiss-Meditec) are often used to report loss of function as negative values (e.g., MD: mean *deviation* of $-10$ dB), while those working with Octopus perimeters (Haag-Streit) often report loss as a positive value (e.g., MD: mean *defect* of 10 dB). Furthermore, clinicians working with Octopus perimeters often use different grids, even though the 24-2 and 10-2 grids are also available. Usually, fields cannot be directly compared because other setting ($L_{max}$) are also different and because different algorithms are used.

@@

For a custom grid of location, it is possible to create and set the location map and its graphical parameters. visualFields has a function `vfgpar()` to generate graphical parameters for a given grid. Here is an example with the Octopus "G1" map, that uses `vfread` to read visualField's ready visual field data.

[I WANTED TO SHOW SOME OF THE FUNCTIONALITY OF THE LOADOCTOPUS FUNCTION, BUT I COULD DOCUMENT THIS ELSEWHERE]

[THE PROBLEM HERE IS THAT NOW PEOPLE WORKING WITH THE OCTOPUS HAVE TO FIND THE COORDINATES ELSEWHERE, AGAIN, BECAUSE THEY DO NOT HAVE ACCESS TO g1.csv. COULD IT BE PROVIDED IN THE PACKAGE?]

```{r}
# read the Octopus G1 coordinates and create the locmap
g1 <- list(coord = read.csv("./g1.csv"), name = "G1", desc = "The G1 location map used by Octopus perimeters", bs = numeric())
str(g1)
# set the G1 locmap
setlocmap(g1)
# create and set the graphical parameters
setgpar(vfgpar(g1$coord))
# read visual fields and plot the sensitivities
vfplot(vfread("./vfg1.csv"), type = "s")
```
